---
title: "Analysis of Personal Movement Using Activity Monitoring Devices"
author: "CM"
date: "9/7/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This project makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. The goal of this project is to write a report that answers the questions detailed below.

##Loading and preprocessing the data 

1. Load the data from the 'activity.csv' into the R programme. 

```{r}
data <- read.csv("activity.csv", header = TRUE, sep = ",", na.strings = "NA")
```

2. Look at the summary and structure of the dataset. 

```{r}
summary(data)
```

```{r}
str(data)
```

3. Look at the first six rows of the dataset 'activity.csv'. 

```{r}
head(data)
```

4. Convert the "date" variable to a "Date" class and the "interval" variable to a "factor".

```{r}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- factor(data$interval)
```

### What is mean total number of steps taken per day? 

1. Subset the dataset to remove the missing values. 

```{r}
NA_index <- is.na(as.character(data$steps))
data_no_NA <- data[!NA_index,]
head(data_no_NA)
```

2. Aggregate the number of steps taken per day. 

```{r}
steps_each_day <- aggregate(steps ~ date, data = data_no_NA, sum)
colnames(steps_each_day) <- c("date", "steps")
```

3. Create a histogram of the total number of steps taken per day. 

```{r}
hist(as.numeric(steps_each_day$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
dev.copy(png, "hist1.png", width=480, height=480)
dev.off()
cat("\014") 
```

4. Calculate and report the mean and median of the total number of steps taken per day. 
```{r}
mean(steps_each_day$steps)
```

```{r}
median(steps_each_day$steps)
```

### What is the average daily activity pattern? 

1. Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis). 

```{r}
steps_per_interval <- aggregate(data_no_NA$steps, by=list(interval=data_no_NA$interval), FUN=mean)
colnames(steps_per_interval) <- c("interval", "average_steps")
plot(as.integer(levels(steps_per_interval$interval)), steps_per_interval$average_steps, type="l",
     xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="blue")
dev.copy(png, "Plot1.png", width=480, height=480)
dev.off()
cat("\014") 
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_steps <- max(steps_per_interval$average_steps)
max_steps
```

```{r}
intervale_max_steps<-steps_per_interval[which.max(steps_per_interval$average_steps),]$interval
intervale_max_steps
```

### Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s).

```{r}
sum(is.na(as.character(data$steps)))
sum(is.na(as.character(data$date)))
sum(is.na(as.character(data$interval)))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}
NA_index <- which(is.na(as.character(data$steps)))
complete_data <- data
complete_data[NA_index, ]$steps<-unlist(lapply(NA_index, FUN=function(NA_index){
                steps_per_interval[data[NA_index,]$interval==steps_per_interval$interval,]$average_steps
                }))
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
summary(complete_data)
```

```{r}
str(complete_data)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
steps_each_day_complete <- aggregate(steps ~ date, data = complete_data, sum)
colnames(steps_each_day_complete) <- c("date", "steps")
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
dev.copy(png, "hist2.png", width=480, height=480)
dev.off()
cat("\014")
```

5. Calculate the mean and median total number of steps taken per day for the complete dataset. 

```{r}
mean(steps_each_day_complete$steps)
```

```{r}
median(steps_each_day_complete$steps)
```

The mean of the complete dataset is equal to the mean of the dataset without missing values. The median of the complete dataset has shifted from 10765 to 10766. Therefore, the mean and median for the complete dataset are almost identical.

### Are there differences in activity patterns between weekdays and weekends?

```{r}
complete_data$day <- as.factor(weekdays(complete_data$date))
complete_data$is_weekday <- ifelse(!(complete_data$day %in% c("Saturday","Sunday")), TRUE, FALSE)
weekdays_data <- complete_data[complete_data$is_weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)
weekends_data <- complete_data[!complete_data$is_weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)
week_data$day <- as.factor(week_data$day)
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, layout = c(1,2), type ="l", ylab="Number of Steps")
dev.copy(png, "Plot3.png", width=480, height=480)
dev.off()
cat("\014")
```
